Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-06-04 _tex/paper2.tex
Built site for gh-pages

Drawing conclusions about the real world from observed data always
requires assumptions, regardless of whether the data are observational
or experimental \citep{Kohler_et_al_2019, Deffner_et_al_2022}. This
study is no exception to this principle. By adopting causal inference as
the theoretical backbone of our proposed framework, we necessarily rely
on expert knowledge and assumptions about the variables' causal
structure that are often untestable at the outset
\citep{Hernan_et_al_2025}. As such, our framework does not yield
automatic answers. Instead, it encourages researchers to formulate
precise questions and to explicitly articulate their assumptions to
develop a generalizable understanding of the CJ system
\citep{Rohrer_et_al_2022, Deffner_et_al_2022, Sterner_et_al_2024}. This
clarity is essential because a model's ability to produce accurate
estimates and valid inferences depends heavily on how well the data and
inferential goals align with its underlying assumptions
\citep{Kohler_et_al_2019}--a property we cannot yet guarantee for our
models. However, this theory-driven approach offers a key benefit: its
lays the foundation for empirically testing the proposed causal
assumptions \citep{Deffner_et_al_2022}.

In light of these limitation, this study presents several challenges
that researchers must navigate--challenges common to many quantitative
studies. These challenges fall into two main categories. The first
concerns the foundational knowledge researchers need to apply our
proposed framework effectively. The second involves the increased
attention researchers must now devote to their specific research
context.

@@

In terms of foundational knowledge, researchers must face two core
challenges to use our framework effectively. First, they need a solid
understanding of causal inference principles. Second, they must learn to
translate the functional and probabilistic content of conceptual models
into bespoke statistical models. One example that illustrates the
importance of a sound knowledge of causal inference is the recurrent
assumption that predictor variables are ``relevant'' to the research
context---interpreting this relevance as their inclusion in a
\emph{sufficient adjustment set}
\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. However, we do
not fully explore what this assumption entails or its implications for
model specification, estimation, and inference. As a result, researchers
must engage deeply with these and other complex ideas, such as how SCMs
encode functional and probabilistic information. To assist this effort,
we include key references throughout the study to guide motivated
readers toward deeper understanding\footnote{refer to footnote 1, 2, and
  the detailed online document referenced in the Declarations section}.

Moreover, developing the skills to translate functional and
probabilistic information into bespoke statistical models has its
challenges. Although we argue that Bayesian inference methods offer a
more accessible path for many applied researchers to develop these
skills--since they reduce the need for specialized knowledge in areas
like optimization theory, which frequentist methods often require--these
methods still demand familiarity with technical concepts, such as
probabilistic programming languages (PPLs), probability distributions,
and convergence, that researchers must acquire. Nevertheless, to support
researchers in building these skills, we include links to model code and
alternative specifications in the \textbf{Declarations} section of this
document.

@@

Regarding attention to the research context, researchers face two
additional challenges. First, they must verify that the conceptual model
provides a \emph{faithful} \footnote{Avoid confusing this term with the
  \emph{faithfulness assumption} described in the causal inference
  literature. For further details, see footnote 1.} representation of
the CJ system under study. Second, they must assess whether the
statistical translation of the model can accurately estimate the
intended estimands (i.e., parameters) from empirical data. To ensure a
faithful representation of the CJ system, we encourage researchers to
treat our conceptual and statistical models as starting points rather
than universal solutions for all CJ designs or datasets. As such,
researchers must tackle the challenge of adapting these models to their
specific research contexts, paying close attention to causal
assumptions, experimental design, and model specification. Nonetheless,
our approach offers a transparent framework that enables researchers to
explicitly articulate any new assumptions and design CJ experiments that
support reliable and valid trait measurement.

>>>>

Drawing conclusions about the real world from observed data always
requires assumptions, regardless of whether the data are observational
or experimental \citep{Kohler_et_al_2019, Deffner_et_al_2022}. This
study is no exception to this principle. By adopting causal inference as
the theoretical backbone of our proposed framework, we necessarily rely
on expert knowledge and assumptions about the variables' causal
structure that are often untestable at the outset
\citep{Hernan_et_al_2025}. As such, our framework does not yield
automatic answers. Instead, it encourages researchers to formulate
precise questions and to explicitly articulate their assumptions to
develop a generalizable understanding of the CJ system
\citep{Rohrer_et_al_2022, Deffner_et_al_2022, Sterner_et_al_2024}. This
clarity is essential because a model's ability to produce accurate
estimates and valid inferences depends heavily on how well the data and
inferential goals align with its underlying assumptions
\citep{Kohler_et_al_2019}--a condition we cannot yet guarantee for our
proposed models. Nevertheless, a key strength of this theory-driven
approach is that it lays the foundation for empirically testing the
proposed causal assumptions \citep{Deffner_et_al_2022}, as discussed in
Section~\ref{sec-discussion_RA}.

This theoretical commitment to causal inference also introduces several
challenges that researchers must navigate--challenges common to many
quantitative studies. These fall into two main categories. The first
concerns the foundational knowledge researchers need to apply our
proposed framework effectively. The second involves the increased
attention researchers must now devote to their specific research
context.

@@

To effectively apply our framework, researchers must tackle two main
challenges related to foundational knowledge. First, they need a solid
understanding of causal inference principles. Second, they must learn to
translate the functional and probabilistic content of conceptual models
into bespoke statistical models. One example that illustrates the
importance of a sound knowledge of causal inference is the recurrent
assumption that predictor variables are ``relevant'' to the research
context--interpreting this relevance as their inclusion in a
\emph{sufficient adjustment set}
\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. However, we do
not fully explore what this assumption entails or its implications for
model specification, estimation, and inference. Consequently,
researchers must deeply engage with these and other complex ideas,
including how SCMs encode functional and probabilistic information. To
assist this effort, we include key references throughout the study to
guide motivated readers toward a deeper understanding of these concepts
\footnote{refer to footnote 1, 2, and the detailed online document
  referenced in the Declarations section}.

Developing the skills to translate conceptual models into bespoke
statistical models also presents challenges. Although Bayesian inference
methods offer a more accessible path for many applied researchers to
develop these skills--by reducing the need for specialized knowledge in
areas like optimization theory, which frequentist methods often
require--they still demand familiarity with other technical concepts
such as probabilistic programming languages (PPLs), probability
distributions, and convergence, which researchers may find hard to
master. Thus, to support this skill development, we include links to the
statistical model code and alternative specifications in the
\textbf{Declarations} section of this document.

@@

Researchers face two additional challenges when applying our framework
to their specific research context. First, they must verify that the
conceptual model provides a \emph{faithful} \footnote{Avoid confusing
  this term with the \emph{faithfulness assumption}
  \citep{Neal_2020, Hernan_et_al_2025} described in the causal inference
  literature. For further details, see footnote 1.} representation of
the CJ system under study. Second, they need to assess whether the
statistical translation of the model can accurately estimate the
intended estimands (i.e., parameters) from empirical data. To ensure a
faithful representation of the CJ system, we encourage researchers to
treat our conceptual and statistical models as starting points rather
than universal solutions for all CJ designs or datasets. This
consideration means researchers must tackle the challenge of adapting
these models to their specific research contexts, paying close attention
to causal assumptions, experimental design, and model specification. Our
approach supports this process by offering a transparent framework that
helps researchers explicitly articulate these new assumptions and
supports the design of CJ experiments that can yield reliable and valid
inferences.
